---
title: "Chapter 6"
author: "Mette Nissen"
date: "12/9/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chapter 6. Analysis of longitudinal data {.tabset}


```{r}
library(dplyr)
library(tidyr)
```

###Load the data sets (BPRS and RATS)

```{r}
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", header = TRUE, sep = " ")
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')
```


###Inspecting the WIDE data

```{r}
names(BPRS)
names(RATS)

summary(BPRS)
summary(RATS)

str(BPRS)
str(RATS)
```

###Treatment weeks seems to be in different colums

###Convert categorical data to factors - BPRS

```{r}
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)

#BPRS to long form
BPRSL <-  BPRS %>% gather(key = weeks, value = bprs, -treatment, -subject)

# Extract the week number
BPRSL <-  BPRSL %>% mutate(week = as.integer(substr(weeks, 5,5)))
```


### A glimpse at the BPRSL data
```{r}
glimpse(BPRSL)
str(BPRS)
```

We now have a dataset with 360 observations and 5 variables -> long form

And the same preparation to RATS

```{r}
# Convert categorical data to factors - RATS
RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)

# Convert data to long form and extract the time

RATSL <- RATS %>%
  gather(key = WD, value = Weight, -ID, -Group) %>%
  mutate(Time = as.integer(substr(WD, 3, 4))) 

# A glimpse at the data
glimpse(RATSL)
str(RATSL)


names(BPRSL)
names(RATSL)
```

In the long form time is in one column and in  wide time variables are in different columns

write.csv(RATSL, "data/rats_long.csv")
write.csv(BPRSL, "data/bprs_long.csv")

```{r}
#Access the package ggplot2
library(ggplot2)
```

##Exploring RATS and RATSL dataset

```{r}
glimpse(RATS)
head(RATS)
str(RATS)
summary(RATS)


glimpse(RATSL)
head(RATSL)
str(RATSL)
summary(RATSL)
```

 From the Wide dataset we can see that we have 3 groups with Group 1 : 8 rats, Group 2 : 4 rats and Group 3 : 4 rats.
 Weight is measured in 11 timepoints 7 WD apart. Overall weight seems to increase over time. 

```{r}
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=8)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))

# Next weight is standardized with SD and same plot is drawn. 

RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate( stdWeight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()
glimpse(RATSL)

ggplot(RATSL, aes(x = Time, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=8)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$stdWeight), max(RATSL$stdWeight)))
```

Drawing the lines of individual rats within different groups over time. On X axis we have timepoints and on Y-axis we have weight in grams. 
 Individuals are marked with a different linetype. Different groups are in separate graphs.  
In this individual graph we can clearly see that the baseline is much lower in group 1. In group 2 is an outlier with higher starting weight which increases the total mean value.

With standardization weight mean is 0 with SD of 1. Standardization is done by grouping by time, so the time variable doesn´t show similar trend as before. 


##Combining individuals to a group mean

```{r}
RATSL2 <- RATSL
RATSL2 <- within(RATSL2, {Time <- factor(Time)})
RATSL2.aov <- aov(Weight ~ Group * Time + Error(ID), data = RATSL2)
summary(RATSL2.aov)

n <- RATSL$Time %>% unique() %>% length()
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()
glimpse(RATSS)
```

### Glimpse the data

```{r}
glimpse(RATSS)
```


### Plotting the mean profiles

```{r}
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1:3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1:3)) +
  #geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(rats) +/- se(rats)")
```

The between groups test indicates that the variable group is significant, 
consequently in the graph we see that the lines for the three groups are rather far apart.
The within subject test indicate that there is a significant time effect, 
in other words, the groups show increase in weight over time. 
The slopes of the lines are increasing in all groups. 
One group starts at a lower mean of weight and increase over time seems to be less than in other two groups.

From the previous graphs we could see already that group 2 has an outlier, but let us see that in better way - boxplot

```{r}
ggplot(RATSL2, aes(x = factor(Time), y = Weight, fill = Group)) + geom_boxplot(position = position_dodge(width = 0.9)) + scale_x_discrete(name = "Time") 

RATSL8S <- RATSL %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise(mean=mean(Weight) ) %>%
  ungroup()
```

Glimpse the data
```{r}
glimpse(RATSL8S)
```


### Drawing a boxplot of the mean versus treatment

```{r}
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Days")
```

From the boxplot we can see the group 2 outlier clearly, but also in groups 1 and 3 are single outliers but they follow means trend. 
For sciences sake let us make these groups smaller and filter outliers.

```{r}
RATSL8S1 <- RATSL8S %>%
  filter(mean < 550)
RATSL8S1 <- RATSL8S %>%
  filter(mean > 240)
RATSL8SG3 <- subset(RATSL8S, Group==3, select=c(Group, ID, mean)) %>%
  filter(mean > 500)
RATSL8SG12 <- subset(RATSL8S1, Group==1:2, select=c(Group, ID, mean))
RATSL8S1 <- rbind(RATSL8SG12, RATSL8SG3)
```

### Drawing a boxplot without outliers of the mean versus Group

```{r}
ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Days")
```


Without outliers groupmeans are very different

##ANOVA

Next We use anova for analysis of variance, t-test canẗ be used because variables aren´t bimodal.

```{r}
fit1 <- lm(mean ~ Group, data = RATSL8S1)
summary(fit1)
anova(fit1)
```

Here we can see the group means signifiqant difference.

Using baseline as a covariate to the model we can see if there is still difference ?

```{r}
baseline <- RATS$WD1
RATSL8S2 <- RATSL8S %>%
  mutate(baseline)
fit2 <- lm(mean ~ baseline + Group, data = RATSL8S2)
summary(fit2)
anova(fit2)
```

Signifiqance was lost. So the baseline should be taken account choosing a treatment group. 